Course Project 1
=================
Brynjólfur Gauti Jónsson

Let's start by loading in the libraries we'll use.
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(ggthemes)
```

# Reading and cleaning the data

### Let's start by reading in the data and looking at it.
It's stored as a csv so reading it in should be very simple. After that we'll use head and str to get a quick look at the data.
```{r reading data}
data <- read.csv('activity.csv')
head(data)
str(data)
```


### Cleaning the data

We see that the date variable is stored as a factor. Let's change it into a date.

```{r variable cleaning}
data$date <- as.Date(data$date, format = '%Y-%m-%d')
str(data)
```

### Plotting the total steps per day:

We'll use ggplot for this as it's very customizable. First we have to create a vector containing the daily total steps.

```{r}
sum <- data %>%
    group_by(date) %>%
    summarise(sum = sum(steps)) %>%
    arrange(desc(sum))
head(sum)
```

```{r histSteos}
gghist <- ggplot(data = sum, aes(sum))
gghist + geom_histogram(na.rm = TRUE, binwidth = 2000) + theme_tufte() + 
    xlab('Steps') + ylab('Days') + ggtitle('Total Step Counts')
```

### Calculating the mean and median of steps per day:
```{r mean and median}
meansteps <- mean(sum$sum, na.rm=TRUE)
mediansteps <- median(sum$sum, na.rm=TRUE)
```

The mean number of steps is: `r format(meansteps, scientific=FALSE)`, The median is: `r mediansteps`.


### Plotting a time series of the steps over intervals

```{r timeSeries}
# Making the ticks for the x-axis
intervaltime <- str_pad(unique(data$interval), 4, pad='0')
intervaltime <- format(strptime(intervaltime, format='%H%M'), format='%H:%M')

g2 <- ggplot(data, aes(x=interval, y=steps))
g2 + stat_summary(fun.y=mean, geom='line', na.rm=TRUE) + 
    xlab('Time of Day')+
    ylab('Mean Number of Steps') + 
    ggtitle('Mean Number of Steps For Different Times of Day') +
    theme_tufte(base_size = 12) + 
    scale_x_continuous(breaks=seq(1, 2355, length.out = 12), 
                       labels=intervaltime[seq(1, 288, by=288/12)])
```

Now to find which 5-minute interval contains the largest average number of steps.

```{r means per interval}
means <- data %>%
    group_by(interval) %>%
    summarise(avgsteps = mean(steps, na.rm=TRUE)) %>%
    arrange(desc(avgsteps))
head(means)
```

The interval `r means[1,1]` contains the highest average of `r means[1,2]`

## Imputing missing values

To see how many NAs are in the data lets first see where the NAs are found:
``r colSums(is.na(data))``
```{r}
colSums(is.na(data))
```

It seems that the only NAs are in the steps category so the number of NAs is just the number of NAs in the steps column: `r colSums(is.na(data))[1]` 

### Function to impute
Now let's find a way to impute the missing values. When we have an NA in the steps variable we'll make it into the mean of steps grouped by the 5 minute interval in which the measurement would have taken place.

```{r imputing data}
impute <- data %>% 
    group_by(interval) %>%
    mutate_at(., 'steps',funs(replace(., which(is.na(.)), mean(.,na.rm=TRUE))))

head(impute, 10)
```


Now let's remake the histogram of total steps only this time with imputed values. First we have to create a new vector of total daily steps.

```{r}
sum2 <- impute %>%
    group_by(date) %>%
    summarise(sum = sum(steps)) %>%
    arrange(desc(sum))
head(sum)
```


```{r histImpute}
gghist2 <- ggplot(data = sum2, aes(sum))
gghist2 + geom_histogram(na.rm = TRUE, binwidth = 2000) + theme_tufte() + 
    xlab('Steps') + ylab('Days') + ggtitle('Total Step Counts')

meanimpute <- mean(sum2$sum, na.rm=TRUE)
medianimpute <- median(sum2$sum, na.rm=TRUE)

```

The mean of imputed data is `r format(meanimpute, scientific=FALSE)` and the median is `r format(medianimpute, scientific=FALSE)`. We see that the median and mean have become the same. The distribution also beacme a little more normalized by imputing it, as one might expect since we were using means to impute.

## Are there differences in activity between weekdays and weekends?

Let's make a new variable for weekday vs weekend and plot the mean number of steps for each, minutewise.
```{r weekdayVsWeekend, fig.width= 12, fig.height= 6}

impute$weekday <- weekdays(impute$date)
impute[impute$weekday %in% c('Monday', 'Tuesday', 'Wednesday', 'Thursday',
                             'Friday'),4] <- 'Weekday'
impute[impute$weekday %in% c('Saturday', 'Sunday'),4] <- 'Weekend'

g2 <- ggplot(impute, aes(x=interval, y=steps))
g2 + stat_summary(fun.y=mean, geom='line', na.rm=TRUE) + xlab('Time of Day') +
    ylab('Mean Number of Steps') + 
    ggtitle('Mean Number of Steps For Different Times of Day') +
    theme_tufte(base_size = 13) +
    scale_x_continuous(breaks=seq(1, 2355, length.out = 12), 
                       labels=intervaltime[seq(1, 288, by=288/12)]) +
    facet_grid(. ~ weekday)
```



